Linear analysis of the vertical shear instability in astrophysical disks

### 2015-01-25 application.tex
almost done with numerical results section

\subsection{Numerical example}
Consider  the MMSN at $r=30\mathrm{AU}$, with $\gamma=1.4$ and
$\mu=2$. Eq. \ref{mmsn_epsilon} gives $\epsilon \simeq
0.058$ which implies, using Eq. \ref{bcrit_mmsn}, a critical 
relaxation timescale of  
\begin{align*}
  \beta_\mathrm{crit} \simeq 0.062. 
\end{align*} 
To estimate a thermal relaxation timescale, we set $\khat=20$ 
in Eq. \ref{beta_mmsn} to obtain 
\begin{align*}
  \beta \simeq 0.042
\end{align*}
at $r=30\mathrm{AU}$. This indicates the disk supports the fundamental
isothermal VSI, but only marginally. Using the above value of $\beta$,
MMSN disk parameters and $\khat=20$ in our linear code, we find a
growth rate of 
\begin{align*}
  \nu = 3.3\times10^{-3}\Omega_k.
\end{align*}
This corresponds to a characteristic growth time of $t_g\sim 50$
orbits. In fact, this mode was found to be the most unstable for
$\khat=20$. 
% and
% physically acceptable -> other modes
% are crazy

We also solved the linear problem with varying $\khat$  
and $\beta$ determined by Eq. \ref{beta_mmsn}. We find the most
unstable wavenumber occurs at $\khat\simeq60$ with $\nu\simeq
3.6\times10^{-3}\Omega_k$, i.e. only slightly faster than that above. 
However, the eigenfunctions for large $\khat$ were found to have large
amplitude oscillations near the vertical boundaries.

 %only find one unstable mode -> close to marginal 
 %larger kx

>>>>

\subsection{Numerical example}
Consider  the MMSN at $r=30\mathrm{AU}$, with $\gamma=1.4$ and
$\mu=2$. Eq. \ref{mmsn_epsilon} gives $\epsilon \simeq
0.058$ which implies, using Eq. \ref{bcrit_mmsn}, a critical 
relaxation timescale of  
\begin{align*}
  \beta_\mathrm{crit} \simeq 0.062. 
\end{align*} 
To estimate a thermal relaxation timescale, we set $\khat=20$ 
in Eq. \ref{beta_mmsn} to obtain 
\begin{align*}
  \beta \simeq 0.042.
\end{align*}
(In fact, Fig. \ref{mmsn_bcrit_bcool} shows that $\beta$  is not
sensitive to $\khat$ at $r=30\mathrm{AU}$, provided that $\khat\gg1$.)
The disk supports the fundamental isothermal VSI, but only
marginally. Using the above value of $\beta$, MMSN disk parameters and
$\khat=20$ in our linear code, we find a  
growth rate of 
\begin{align*}
  \nu = 3.3\times10^{-3}\Omega_k.
\end{align*}
This corresponds to a characteristic growth time of $t_g\sim 50$
orbits. This mode was found to be the most unstable for
$\khat=20$. 
% and
% physically acceptable -> other modes
% are crazy

We also solved the linear problem with varying $\khat$  
and $\beta$ determined by Eq. \ref{beta_mmsn}. We find the most
unstable wavenumber occurs at $\khat\simeq60$ with $\nu\simeq
3.6\times10^{-3}\Omega_k$, i.e. only slightly faster than $\khat=20$.   
However, the eigenfunctions for large $\khat$ were found to have large
amplitude oscillations near the vertical boundaries. This is shown in
Fig. \ref{mmsn_eigenW}. These modes may not be important since they
mostly affect the disk atmosphere where there is little mass.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/eigenvectorW_mmsnkx20.ps}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 0cm]{figures/eigenvectorW_mmsnkx60.ps}
  \caption{Most unstable VSI modes in the MMSN at
    $r=30\mathrm{AU}$. The eigenfunction $W$ for $\khat=20$ (top) and
    $\khat=60$ (bottom) are shown. The real (imaginary) part of $W$
    is plotted as the solid (dotted) lines.\label{mmsn_eigenW}}  
\end{figure}

% only find one unstable mode -> close to marginal 
% larger kx

### 2015-01-25 numerical.tex
almost done with numerical results section

\subsection{Vertically non-isothermal disks}
%gamma = Gamma = 1.4, vary bcool

@@

>>>>

\subsection{Vertically non-isothermal disks}
We briefly consider vertically non-isothermal disks with
$\Gamma=1.4$. Other disk parameters are $p=0$, $q=-1.0$ and $\epsilon
= 0.05$. These are again chosen for comparison with \cite{nelson13}. 
In this case we set $\zmax\simeq2.2\epsilon r$, which is close to the
zero-density surface.

\subsubsection{$N_z^2\equiv 0$}

Consider first the case $\gamma=1.4$ so that there is no vertical
entropy gradient. We plot in Fig. the growth rates for the
fundamental VSI for $\beta\in[10^{-2},10^{2}]$. We find growth rates
do not reach zero, implying the VSI operates regardless of the thermal
relaxation timescale if $\gamma=\Gamma$. This result is consistent
with \cite{nelson13}.

We also note that growth rates for $\beta = 10^{-2}$ and
$\beta=10^{2}$ behave similarly. This is not surprising because
the linearized energy equation (Eq. \ref{lin_energy}) implies that 
\begin{align}
  W = \frac{\left(\Gamma^{-1} -
      \ii\hat{\sigma}\beta\right)}{\left(1-\ii\hat{\sigma}\beta\right)}
  Q \text{ for } \gamma\equiv \Gamma.   
\end{align}
Thus, $W \to Q/\Gamma$ as $\beta\to0$ and $W\to Q$ as
$\beta\to\infty$. Since $\Gamma=O(1)$, these two limits are expected
to produce qualitatively similar results.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/growth_vnoniso1} 
  \caption{Numerically-determined growth rates of the fundamental VSI
    mode in neutrally-buoyant, vertically non-isothermal disks
    ($\Gamma=\gamma=1.4$), with $p=0$, $q=-1$ and  $\epsilon=0.05$, for
    a range of thermal relaxation timescales
    $\beta$. \label{growth_vnoniso1}}     
\end{figure}

Interestingly, Fig. \ref{growth_vnoniso1} show that growth rates
maximize for $\beta=O(1)$. However, all growth rates are similar to
order-of-magnitude for $0.1\lesssim\khat\lesssim 100$.

\subsubsection{$N_z^2\geq0$}

@@

Finally, we introduce buoyancy in a vertically non-isothermal disk by
setting $\gamma=2$. This situation was not simulated in 
\cite{nelson13}, but is similar in spirit with vertically isothermal 
disks evolved with $\gamma>1$, as considered in the previous section.

Fig. plots the growth rates for this case. The trend is
qualitatively similar to that for vertically isothermal disks with
$N_z^2\geq 0$ (Fig. \ref{relax_growth_num}), namely growth rates reach
zero at finite $\khat$.
